A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-02-10 paper4.Rmd
add interaction checks for postneonatal ...

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[tablesonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, cache= TRUE) ## NOTE!!
knitr::opts_knit$set(global.par = TRUE)
##
library(knitr)
library(eha)
starttid <- proc.time()
```

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
##infdat <- readRDS("data/infdat4.rds") Already done
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
par(cex.axis = 0.7)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR", lwd = 1.5, cex = 0.7)
abline(h = c(0.05, 0.1, 0.15), lty = 3)
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3}
oldpar <- par(mfrow = c(1, 3))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
umeanavg <- umetemp$emeantemp[1:52]
##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, umaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
m12 <- c(-30, -15, 0, 14.5, 25)
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, umeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
bmeanavg <- bjurtemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
## Inland:
bmaxavg <- stentemp$emaxtemp[1:52]
bminavg <- stentemp$emintemp[1:52]
bmeanavg <- stentemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Inland", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
par(oldpar)
```

@@

```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
bmtemp <- rbind(umetemp, bjurtemp, stentemp)
umt <- with(bmtemp, tapply(meantemp, year, mean))
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
abline(h = 0)
abline(h = 1:6, lty = 3)
```

@@

```{r managedata}
library(survival)
neo <- age.window(infdat, c(0, 28 / 365))
neo$ageIvl <- "neo"
pneo <- age.window(infdat, c(28/365, 1))
pneo$ageIvl <- "postneo"
all <- rbind(neo, pneo)
all$ageIvl <- factor(all$ageIvl)
##all$enter <- round(all$enter * 365)
##all$exit <- round(all$exit * 365)
##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
  ##               data = all)
##dr.all <- drop1(fit.all, test = "Chisq")

inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
                     extemp + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
                 cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
tabneo <- inftab[inftab$ageIvl == "neo", ]
tabpost <- inftab[inftab$ageIvl == "postneo", ]
tabpost$age <- factor(tabpost$age)
tabneo$age <- factor(tabneo$age)
```

```{r joint}
fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
                       (extemp + emeantemp + lowTemp + highTemp + socBranch + period), 
                   data = inftab, time = age)
dr.all <- drop1(fit.all, test = "Chisq")
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabpost, time = age))
```

@@

We can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects.

@@

```{r formneo}
formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + extemp + emeantemp + socBranch + period +
                     subreg + sex + parity")
```

@@

```{r neoanalwp, results = 'asis'}
##library(survival)
winter <- tabneo[!tabneo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.neow <- tpchreg(formneo, data = winter, time = age)
dr <- drop1(fit.neow, test = "Chisq")
```

```{r neoanalw, results = "asis", cache = FALSE}
ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanalsp}
##library(survival)
#summer <- neo[neo$summer, ]
##summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.neos <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
dr.neos <- drop1(fit.neos, test = "Chisq")
```

```{r neoanals, results = 'asis', cache = FALSE}
ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
     label = "tab:neoanals", keep = NULL)
```

@@

@@

```{r postanalwp, results='asis'}
##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
  ##              + subreg + period, data = infdat, cuts = c(0, 1))
##winter <- gtpneo[gtpneo$summer == 0, ]
##winter$season <- factor(winter$season)

fit.postw <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + extemp + emeantemp), 
                     data = tabpost[!tabpost$summer, ], time = age)
dr.postw <- drop1(fit.postw, test = "Chisq")
```

@@

This shows that the temperature effects do not vary with time period,
but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
to make sure, will be joined when appropriate.

```{r setformw}
form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + subreg + sex + parity")
```

@@

```{r postw1, results = "asis", cache = FALSE}
ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, parity, and subregion.", label = "tab:postw1", keep = 5)
```

Table \@ref(tab:postw1) shows that

@@

Table \@ref(tab:postw2) shows that

@@

```{r postw3p, results = "asis"}
fit.postw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
dr.postw3 <- drop1(fit.postw3, test = "Chisq")
```

@@

@@

```{r postanalsp, results='asis'}
##summer <- gtpneo[gtpneo$summer == 1, ]
##winter$season <- factor(winter$season)

fit.posts <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + period + 
                        subreg + parity + sex, 
                    data = tabpost[tabpost$summer, ], time = age)
dr <- drop1(fit.posts, test = "Chisq")
```

>>>>

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
#- \usepackage[tablesonly]{endfloat}
#- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA) ## NOTE!!
knitr::opts_knit$set(global.par = TRUE)
##
library(knitr)
library(eha)
starttid <- proc.time()
```

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
##infdat <- readRDS("data/infdat4.rds") Already done
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
par(cex.axis = 0.7)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR", lwd = 1.5, cex = 0.7)
abline(h = c(0.05, 0.1, 0.15), lty = 3)
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3, cache = TRUE}
oldpar <- par(mfrow = c(1, 3))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
umeanavg <- umetemp$emeantemp[1:52]
##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, umaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
m12 <- c(-30, -15, 0, 14.5, 25)
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, umeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
bmeanavg <- bjurtemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
## Inland:
bmaxavg <- stentemp$emaxtemp[1:52]
bminavg <- stentemp$emintemp[1:52]
bmeanavg <- stentemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Inland", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
par(oldpar)
```

@@

```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå.", cache = TRUE}
bmtemp <- rbind(umetemp, bjurtemp, stentemp)
umt <- with(bmtemp, tapply(meantemp, year, mean))
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
abline(h = 0)
abline(h = 1:6, lty = 3)
```

@@

```{r managedata}
if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
    neo <- age.window(infdat, c(0, 28 / 365))
    neo$ageIvl <- "neo"
    pneo <- age.window(infdat, c(28/365, 1))
    pneo$ageIvl <- "postneo"
    all <- rbind(neo, pneo)
    all$ageIvl <- factor(all$ageIvl)
    ##all$enter <- round(all$enter * 365)
    ##all$exit <- round(all$exit * 365)
    ##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
    ##               data = all)
    ##dr.all <- drop1(fit.all, test = "Chisq")

    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
                         extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
                     cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
    tabneo <- inftab[inftab$ageIvl == "neo", ]
    tabpost <- inftab[inftab$ageIvl == "postneo", ]
    tabpost$age <- factor(tabpost$age)
    tabneo$age <- factor(tabneo$age)
    saveRDS(tabneo, file = "data/tabneo.rds")
    saveRDS(tabpost, file = "data/tabpost.rds")
    saveRDS(inftab, file = "data/inftab.rds")
}else{
    inftab <- readRDS("data/inftab.rds") #The rest are read later...
}
```

```{r joint, cache = TRUE}
fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
                       (extemp + extemp.1 + emeantemp + socBranch + period), 
                   data = inftab, time = age)
dr.all <- drop1(fit.all, test = "Chisq")
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
  ##               data = tabpost, time = age))
```

@@

We can see that the variables *emeantemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects.

@@

```{r formneo}
formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socBranch + period +
                     subreg + sex + parity")
tabneo <- readRDS("data/tabneo.rds")
```

@@

```{r neoanalwp, results = 'asis', cache = TRUE}

##library(survival)
winter <- tabneo[!tabneo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.neow <- tpchreg(formneo, data = winter, time = age)
dr <- drop1(fit.neow, test = "Chisq")
```

```{r neoanalw, results = "asis", cache = FALSE}
ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
     label = "tab:neoanalw", keep = 6)
```

@@

```{r neoanalsp, cache = TRUE}
##library(survival)
#summer <- neo[neo$summer, ]
##summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.neos <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
dr.neos <- drop1(fit.neos, test = "Chisq")
```

```{r neoanals, results = 'asis', cache = FALSE}
ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
     label = "tab:neoanals", keep = 6)
```

@@

```{r setformw}
form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + subreg + sex + parity")
form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socBranch + subreg + sex + parity")
form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socBranch + sex + parity") 
tabpost <- readRDS("data/tabpost.rds")
```

@@

```{r postanalwp, results='asis', cache = TRUE}
##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
  ##              + subreg + period, data = infdat, cuts = c(0, 1))
##winter <- gtpneo[gtpneo$summer == 0, ]
##winter$season <- factor(winter$season)

fit.postw <- tpchreg(oe(event, exposure) ~ period * (extemp + extemp.1 + emeantemp), 
                     data = tabpost[!tabpost$summer, ], time = age)
dr.postw <- drop1(fit.postw, test = "Chisq")
```

@@

This shows that the temperature effects do not vary much with time period,
but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
to make sure, will be joined when appropriate.

@@

```{r postw1, results = "asis", cache = FALSE}
ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for sex and parity.", 
     label = "tab:postw1", keep = 5)
```

Table \@ref(tab:postw1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
(remember that this concerns winter conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
but in the expected direction.

The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
the regions regarding *climate*:

```{r postw1srpu, cache = TRUE}
fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg == "ume", ], time = age)
dr.postw1sru <- drop1(fit.postw1sru, test = "Chisq")
```

```{r postw1sru, results = "asis"}
ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå, October to March 1895-1913.",
     label = "tab:postw1sru", keep = 4)
```

```{r postw1srps, cache = TRUE}
fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg != "ume", ], time = age)
dr.postw1srs <- drop1(fit.postw1srs, test = "Chisq")
```

```{r postw1srs, results = "asis"}
ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
     label = "tab:postw1srs", keep = 4)
```

We can conclude from Tables \@ref(tab:postw1sru) and \@ref(tab:postw1srs) that the main difference between Umeå and Skellefteå is that
*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.

@@

Table \@ref(tab:postw2) shows that the weather effect is lagged one week (*extemp.1*), and that climate continues to be very important.
Let's look at regional differences.

```{r postw2srpu, cache = TRUE}
fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg == "ume", ], time = age)
dr.postw2sru <- drop1(fit.postw2sru, test = "Chisq")
```

```{r postw2sru, results = "asis"}
ltx2(fit.postw2sru, dr = dr.postw2sru, caption = "Postneonatal mortality, Umeå, October to March 1914-1934.",
     label = "tab:postw2sru", keep = 4)
```

```{r postw2srps, cache = TRUE}
fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg != "ume", ], time = age)
dr.postw2srs <- drop1(fit.postw2srs, test = "Chisq")
```

```{r postw2srs, results = "asis"}
ltx2(fit.postw2srs, dr = dr.postw2srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1914-1934.", 
     label = "tab:postw2srs", keep = 4)
```

@@

```{r postw3p, results = "asis"}
fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socBranch + subreg + sex + parity, 
                      data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
##fit.postw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
dr.postw3 <- drop1(fit.postw3, test = "Chisq")
```

@@

```{r postw3srpu, cache = TRUE}
fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg == "ume", ], time = age)
dr.postw3sru <- drop1(fit.postw3sru, test = "Chisq")
```

```{r postw3sru, results = "asis"}
ltx2(fit.postw3sru, dr = dr.postw3sru, caption = "Postneonatal mortality, Umeå, October to March 1935-1950.",
     label = "tab:postw3sru", keep = 4)
```

```{r postw3srps, cache = TRUE}
fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg != "ume", ], time = age)
dr.postw3srs <- drop1(fit.postw3srs, test = "Chisq")
```

```{r postw3srs, results = "asis"}
ltx2(fit.postw3srs, dr = dr.postw3srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
     label = "tab:postw3srs", keep = 4)
```

@@

```{r postanalsp, results='asis', cache = TRUE}
##summer <- gtpneo[gtpneo$summer == 1, ]
##winter$season <- factor(winter$season)

fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + period + 
                        subreg + parity + sex, 
                    data = tabpost[tabpost$summer, ], time = age)
dr <- drop1(fit.posts, test = "Chisq")
```
